Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-14 talk/pepm2011/paper.tex
remove a number of "thus"s. simplify two paragraphs in the background section.

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT works by observing the running
program and recording its hot spots into linear execution traces, which are then turned into
machine code. One reason for the popularity of tracing JITs is their relative
simplicity. They can often be added to an interpreter and a lot of the
infrastructure of the interpreter can be reused. They give some important
optimizations like inlining and constant-folding for free. A tracing JIT always
produces linear pieces of code, which simplifies many algorithms that are usually
hard in a compiler, such as register allocation.

@@

The work described in this paper was done in the context of the PyPy project\footnote{\texttt{http://pypy.org}}
\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
be implemented in a simple yet efficient way. 
When implementing a language with PyPy one writes an \emph{interpreter}
for the language in
\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
is a subset of Python chosen in such a way that type inference becomes
possible. The language interpreter can then be compiled (``translated'') with
PyPy's tools into a VM on the C level. Because the interpreter is written at a
relatively high level, the language implementation is kept free of low-level
details, such as object layout, garbage collection or memory model. Those
aspects of the final VM are woven into the generated code during the translation
to C.

@@

The feature that makes PyPy more than a compiler with a runtime system is its
support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
the translation to C, PyPy's tools can generate a just-in-time compiler for the
language that the interpreter is implementing. This process is mostly
automatic; it only needs to be guided by the language implementer using a small number of
source-code hints. Mostly-automatically generating a JIT compiler has many advantages
over writing one manually, which is an error-prone and tedious process.
By construction, the generated JIT has the same semantics as the interpreter.
Many optimizations can benefit all languages implemented as an interpreter in RPython.

Moreover, thanks to the internal design of the JIT generator, it is very easy
to add new \emph{backends} for producing the actual machine code, in addition
to the original backend for the Intel \emph{x86} architecture.  Examples of
additional JIT backends are the one for Intel \emph{x86-64} and an
experimental one for the CLI .NET Virtual Machine \cite{cuni_high_2010}.
PyPy's JIT generator generates a \emph{tracing JIT
compiler}, a concept which we now explain in more details.

@@

Using these classes to implement arithmetic shows the basic problem that a
dynamic language implementation has. All the numbers are instances of either
\lstinline{BoxedInteger} or \lstinline{BoxedFloat}, thus they consume space on the
heap. Performing many arithmetic operations produces lots of garbage quickly,
thus putting pressure on the garbage collector. Using double dispatching to
implement the numeric tower needs two method calls per arithmetic operation,
which is costly due to the method dispatch.

@@

In the example from last section, the following operations in the upper half
of Fig.~\ref{fig:unopt-trace} would produce two
static objects, and would thus be completely removed from the optimized trace:

@@

First, the \lstinline{guard_class} operations can be removed, because the classes of $p_{5}$ and
$p_{6}$ are known to be \lstinline{BoxedInteger}. Second, the \lstinline{get} operations can be removed
and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. Thus the only
remaining operation in the optimized trace would be:

@@

Figure~\ref{fig:semantics} shows the operational semantics for traces. The
interpreter formalized there executes one operation at a time. Its state is
represented by an environment $E$ and a heap $H$, which are potentially changed by the
execution of an operation. The environment is a partial function from variables
to locations and the heap is a partial function from locations to objects. Note
that a variable can never be null in the environment, otherwise the trace would
be malformed. The environment could not directly map variables to object,
because several variables can contain a pointer to the \emph{same} object. Thus
the "indirection" is needed to express sharing.

@@

The state of the optimizer is stored in an environment $E$ and a \emph{static
heap} $S$. The environment is a partial function from variables in the
unoptimized trace $V$ to variables in the optimized trace $V^*$ (which are
themselves written with a
$\ ^*$ for clarity). The reason for introducing new variables in the optimized
trace is that several variables that appear in the unoptimized trace can turn
into the same variables in the optimized trace. Thus the environment of the
optimizer serves a function similar to that of the environment in the semantics: sharing.

>>>>

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT works by observing the running
program and recording its hot spots into linear execution traces. Working on
traces is the central idea of a tracing JIT. Those traces are optimized and
turned into machine code.

One reason for the popularity of tracing JITs is their relative
simplicity. They can often be added to an interpreter and a lot of the
infrastructure of the interpreter can be reused. They give some important
optimizations like inlining and constant-folding for free. A tracing JIT always
produces linear pieces of code, which simplifies many algorithms that are usually
hard in a compiler, such as register allocation.

@@

The work described in this paper was done in the context of the PyPy project\footnote{\texttt{http://pypy.org}}
\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
be implemented in a simple yet efficient way. 
When implementing a language with PyPy one writes an \emph{interpreter}
for the language in
\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
is a subset of Python chosen in such a way that type inference becomes
possible. The language interpreter can then be compiled (``translated'') with
PyPy's tools into a VM on the C level. During translation to C, many low-level
aspects of the final VM, such as object layout, garbage collection and memory
model, are woven into the generated code. Therefore the interpreter itself can
remain at a relatively high level of abstraction.

@@

The feature that makes PyPy more than a compiler with a runtime system is its
support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
the translation to C, PyPy's tools can generate a tracing just-in-time compiler for the
language that the interpreter is implementing. This process is mostly
automatic; it only needs to be guided by the language implementer using a small number of
source-code hints. Mostly-automatically generating a JIT compiler has many advantages
over writing one manually, which is an error-prone and tedious process.
By construction, the generated JIT has the same semantics as the interpreter.
Many optimizations can benefit all languages implemented as an interpreter in RPython.

Moreover, thanks to the internal design of the JIT generator, it is very easy
to add new \emph{backends} for producing the actual machine code.  Examples of
JIT backends that are implemented are the one for Intel x86 and x86-64 and an
experimental one for the CLI .NET Virtual Machine \cite{cuni_high_2010}.

@@

Using these classes to implement arithmetic shows the basic problem that a
dynamic language implementation has. All the numbers are instances of either
\lstinline{BoxedInteger} or \lstinline{BoxedFloat}, therefore they consume space on the
heap. Performing many arithmetic operations produces lots of garbage quickly,
which puts pressure on the garbage collector. Using double dispatching to
implement the numeric tower needs two method calls per arithmetic operation,
which is costly due to the method dispatch.

@@

In the example from last section, the following operations in the upper half
of Fig.~\ref{fig:unopt-trace} produce two
static objects, and can be completely removed from the optimized trace:

@@

First, the \lstinline{guard_class} operations can be removed, because the classes of $p_{5}$ and
$p_{6}$ are known to be \lstinline{BoxedInteger}. Second, the \lstinline{get} operations can be removed
and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. The only
remaining operation in the optimized trace would be:

@@

Figure~\ref{fig:semantics} shows the operational semantics for traces. The
interpreter formalized there executes one operation at a time. Its state is
represented by an environment $E$ and a heap $H$, which are potentially changed by the
execution of an operation. The environment is a partial function from variables
to locations and the heap is a partial function from locations to objects. Note
that a variable can never be null in the environment, otherwise the trace would
be malformed. The environment could not directly map variables to object,
because several variables can contain a pointer to the \emph{same} object. 
The "indirection" is needed to express sharing.

@@

The state of the optimizer is stored in an environment $E$ and a \emph{static
heap} $S$. The environment is a partial function from variables in the
unoptimized trace $V$ to variables in the optimized trace $V^*$ (which are
themselves written with a
$\ ^*$ for clarity). The reason for introducing new variables in the optimized
trace is that several variables that appear in the unoptimized trace can turn
into the same variables in the optimized trace. The environment of the
optimizer serves a function similar to that of the environment in the semantics: sharing.
